---
title: "California Housing Price in 1990"
author:
  - "Phuc Lu"
  - "Meghna Chandrasekar"
  - "Sophia Li"
  - "Youngju Kwon"
output:
  pdf_document:
    extra_dependencies: ["float"]
    latex_engine: xelatex
  html_document: default
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: preamble-latex.tex
spacing: single
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning=FALSE, message=FALSE}
library(readr) # So we can read csv file
library(skimr)
# library(mapview)
# library(sf)
library(dplyr)
library(ggplot2)
options(scipen=999) # Undo Scientific notation

```

[Source](https://www.kaggle.com/datasets/camnugent/california-housing-prices/data) from Kaggle.

```{r}
set.seed(1)
housing_base <- read_csv('housing.csv',show_col_types = FALSE)
housing <- sample(1:length(housing_base$households), 600,replace = FALSE)
housing <- housing_base[housing, ]

## Converting median housing age into categorical data
## Categories: new house, moderately old house, new house
## Intervals 20 years apart from 0 - 60

age = housing$housing_median_age

for (i in 1:length(age)){
    if (age[i] <= 20){
    housing$housing_median_age[[i]] <- "NEW"
  }
  else if (age[i] > 20 && age[i] <= 40){
    housing$housing_median_age[[i]] <- "MODERATE"
  }
  else if (age[i] > 40 && age[i] <= 60){
    housing$housing_median_age[[i]]<-"OLD"
  }
}
```

```{r, eval = FALSE}
housing_sf <- st_as_sf(housing, coords = c("longitude", "latitude"), crs = 4326)
# mapview(housing_sf)
```

```{r}
# Median House Value and Median Income

# Median House Value within the block
#  measured in US Dollars

# Median income of people in the block
# Measured in tens of thousand of US dollars

# Blocks are defined by its longitude and latitude

```



## Our Simple Linear Regression Model is given by
$$
(\text{Median Income})_i = \beta_0 + \beta_1\text{(Median House Value)}_i + \epsilon_i
$$
where $\beta_0$ and $\beta_1$ are unknown constants and $\epsilon_i$ is the random error.

## Hypothesis
We want to know if there is a linear relationship between the median value of houses in a particular block in California in the 1990s and the people living in the block's median income. In other words, we want to test to see if the value of people's house has any linear relationship to their income.\
$$
H_0: \beta_1 = 0 \text{ vs. } H_a: \beta_1 \neq 0
$$

## Plotting Variables
```{r, fig.pos='H', fig.width = 5, fig.height = 3, fig.cap='This graph displays the linear relationship between the median value of a house and the median income of the people living in a particular block in California. We can observe that this relationship is close when in low income area and with cheaper but this relationship is more varied as the median price of houses and people\'s income rises.'}

model <- lm(median_income ~ median_house_value, data = housing)
p <- ggplot(housing, aes(x = median_house_value, y = median_income)) + 
  geom_point() + 
  labs(x = 'Median house value of a block (USD)',
       y = 'Median house income (USD)') +xlim(0, (max(housing$median_house_value+ 20000))) + ggtitle ("Linear Model of Median Income and Median House Value")

p + geom_smooth(method = 'lm',
  formula = 'y ~ x',
  se = F) + theme(plot.title = element_text(size=12))
```

## Checking Assumptions
```{r, fig.pos='H',fig.align='center', fig.width = 7, fig.height = 4, fig.cap= "These graphs show the fan shape which means that as we move go from cheaper to more expensive houses, the variation increases. This fails the assumption of constant variance. As a consequence, we can expect the estimate of the variance, or MSE to be overestimate or underestimate some values of the predictors."}
# Residual Vs. Fitted Plot
model <- lm(median_income ~ median_house_value, data = housing)
house_val <- housing$median_house_value
inc <- housing$median_income

residuals <- model$residuals
par(mfrow=c(1, 2))

plot(fitted(model),residuals, 
     pch = 19,
     main = "Residual vs. Fitted Model",
     xlab = "Residuals",
     ylab = "Fitted Model",
     cex.main = 1)
abline(0,0)

res1 <- model$residuals
plot(residuals, house_val, 
     pch = 19,
     main = "Residual vs. Predictor",
     xlab = "Residuals",
     ylab = "Median House Value",
     cex.main = 1)

```

## Normality
```{r,fig.pos='H',fig.align='center', fig.width = 7, fig.height = 4, fig.cap = "The median house value is distributed approximately Normal"}

par(mfrow = c(1,2))
qqnorm(house_val, pch = 19,
       main = "Normaly Q-Q Plot of Median House Value",
       cex.main = .8)
qqline(house_val)

qqnorm(inc, pch = 19,
       main = "Normaly Q-Q Plot of Median Income",
       cex.main = 1
)
qqline(inc)
```


# Hypothesis Testing
```{r}
## Manual Calculation

alpha <- 0.05

house_val <- housing$median_house_value
inc <- housing$median_income

if (length(house_val) == length(inc)){
  n <- length(house_val)
}else{
  print(paste("Inconsistent n"))
}

# Sxy and Sxx
Sxy <- sum((house_val - mean(house_val)) * (inc - mean(inc)))
Sxx <- sum((house_val - mean(house_val))^2)

# beta 1 and beta 0
beta_hat_1 <- Sxy/Sxx
beta_hat_0 <- mean(inc) - (beta_hat_1 * mean(house_val))

# prediction and residual
y_hat <- beta_hat_0 + (beta_hat_1 * house_val)
residuals <- inc - y_hat

# Mean squared residuals
SSres <- sum(residuals^2)
ms_res <- SSres/ (n - 2)

# Standard Error of beta 1
se_beta_hat_1 <- sqrt(ms_res / Sxx)

# t-statistic
t0 <- beta_hat_1 / se_beta_hat_1

t_quantile <- qt(p = 1 - alpha/2, df = n - 2)

# p-value
p_value <- 2 * pt(abs(t0), df = n - 2, lower.tail = FALSE)

# list(beta_hat_0 = beta_hat_0, beta_hat_1 = beta_hat_1, t0 = t0, t_quantile = t_quantile)
```


```{r}
# print(paste('Sxx: ', Sxx))
# print(paste('beta hat 0: ', beta_hat_0))
# print(paste('beta hat 1: ', beta_hat_1))
# print(paste('sigma hat^2: ', ms_res_house))
# print(paste('t0: ', t0))
# print(paste('t-distribution: ', t_quantile))
# print(paste('p_value: ', p_value))
# print(paste('standard error: ', se_beta_hat_1))
```

Reject $H_0$ if $|t_0| > t_{0.0025, 598}$\
We found that $t_0 = 24.5834068785213$\
So, $|24.5834068785213| >  1.9639388980555$\

```{r}
# print(paste('At the 0.05 significant level, we fail to reject H0. This means that given a particular block in California, there is a linear relation between the median house value and the median income of the people living in the block.'))
```
# Conclusion:
At the 0.05 significant level, we reject H0. This means that given a particular block in California in the 1990s, there is a linear relation between the median house value and the median income of the people living in the block.\




## Confidence Interval
```{r}
model <- lm(median_income ~ median_house_value, data = housing)
model_summary <- summary(model)
beta_hat_0 <- model$coefficients[[1]]
beta_hat_1 <- model$coefficients[[2]]

t_0 <- model_summary$coefficients["median_house_value", "t value"]
p_value <- model_summary$coefficients["median_house_value" , "Pr(>|t|)"]
se <- model_summary$coefficients["median_house_value" , "Std. Error"]

lower <- beta_hat_1 - t_quantile * se_beta_hat_1
upper <- beta_hat_1 + t_quantile * se_beta_hat_1
```
With 95% confidence level, the estimated true median house value of a housing block in California 1990s is said to be between 0.0000107653771071449  and 0.000012634788259628.

# More Confidence Interval
```{r}

```


The predicted median house value of a typical block in California in the 1990s is $\$(3.851572 * 10,000)$ or $\$3,8515.72$




