---
title: "California Housing Price in 1990"
author:
  - "Phuc Lu"
  - "Meghna Chandrasekar"
  - "Sophia Li"
  - "Youngju Kwon"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: preamble-latex.tex
spacing: single
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning=FALSE, message=FALSE}
library(readr) # So we can read csv file
library(skimr)
# library(mapview)
# library(sf)
library(dplyr)
library(ggplot2)
options(scipen=999) # Undo Scientific notation

```

[Source](https://www.kaggle.com/datasets/camnugent/california-housing-prices/data) from Kaggle.

```{r}
set.seed(1)
housing_base <- read_csv('housing.csv',show_col_types = FALSE)
housing <- sample(1:length(housing_base$households), 600,replace = FALSE)
housing <- housing_base[housing, ]

## Converting median housing age into categorical data
## Categories: new house, moderately old house, new house
## Intervals 20 years apart from 0 - 60

age = housing$housing_median_age

for (i in 1:length(age)){
    if (age[i] <= 20){
    housing$housing_median_age[[i]] <- "NEW"
  }
  else if (age[i] > 20 && age[i] <= 40){
    housing$housing_median_age[[i]] <- "MODERATE"
  }
  else if (age[i] > 40 && age[i] <= 60){
    housing$housing_median_age[[i]]<-"OLD"
  }
}
```

```{r, eval = FALSE}
housing_sf <- st_as_sf(housing, coords = c("longitude", "latitude"), crs = 4326)
# mapview(housing_sf)
```

```{r}
# Median House Value and Median Income

# Median House Value within the block
#  measured in US Dollars

# Median income of people in the block
# Measured in tens of thousand of US dollars

# Blocks are defined by its longitude and latitude

# head(housing)

# skim(housing) #%>%
#  select(skim_type, skim_variable, numeric.mean, numeric.sd, numeric.p0, numeric.p50, numeric.p100, skim_hist)
```


```{r}
alpha <- 0.05

x_house_val <- housing$median_house_value
y_inc <- housing$median_income

if (length(x_house_val) == length(y_inc)){
  n_house <- length(x_house_val)
}else{
  n_house <- "Inconsistent n"
}

# Sxy and Sxx
Sxy <- sum((x_house_val - mean(x_house_val)) * (y_inc - mean(y_inc)))
Sxx <- sum((x_house_val - mean(x_house_val))^2)

# beta 1 and beta 0
beta_hat_1 <- Sxy/Sxx
beta_hat_0 <- mean(y_inc) - (beta_hat_1 * mean(x_house_val))

# prediction and residual
y_hat_house <- beta_hat_0 + (beta_hat_1 * x_house_val)
residuals <- y_inc - y_hat_house

# Mean squared residuals
SSres <- sum(residuals^2)
ms_res_house <- SSres/ (n_house - 2)

# Standard Error of beta 1
se_beta_hat_1 <- sqrt(ms_res_house / Sxx)

# t-statistic
t0 <- beta_hat_1 / se_beta_hat_1

t_quantile <- qt(p = 1 - alpha/2, df = n_house - 2)

# p-value
p_value <- 2 * pt(abs(t0), df = n_house - 2, lower.tail = FALSE)

# list(beta_hat_0 = beta_hat_0, beta_hat_1 = beta_hat_1, t0 = t0, t_quantile = t_quantile)
```

## Simple Regression Model
$$
\text{medianIncome}_i = \beta_0 + \beta_1{medianHouseValue}_i
$$

# Hypothesis
We want to know if there is a linear relationship between the median value of houses in a particular block in California and the people's median income.

$H_0: \beta_1 = 0$ vs. $H_a: \beta_1 \neq 0$

```{r}
# print(paste('Sxx: ', Sxx))
# print(paste('beta hat 0: ', beta_hat_0))
# print(paste('beta hat 1: ', beta_hat_1))
# print(paste('sigma hat^2: ', ms_res_house))
# print(paste('t0: ', t0))
# print(paste('t-distribution: ', t_quantile))
# print(paste('p_value: ', p_value))
# print(paste('standard error: ', se_beta_hat_1))
```
$\hat\beta_0 =  1.43434052481747$, $\hat\beta_1 =  0.0000117000826833864$\
$|t_0| =  |24.5834068785213| > t_{0.0025, 598} =  1.9639388980555$
```{r}
# print(paste('At the 0.05 significant level, we fail to reject H0. This means that given a particular block in California, there is a linear relation between the median house value and the median income of the people living in the block.'))
```
# Conclusion:
At the 0.05 significant level, we reject H0. This means that given a particular block in California in the 1990s, there is a linear relation between the median house value and the median income of the people living in the block.\


# Checking Assumptions
```{r, fig.cap="This shows the fan shape which means that as we move left to right, the variation in the data increases. This fails the assumption of constant variance."}
# Residual Vs. Fitted Plot
model <- lm(median_income ~ median_house_value, data = housing)
residuals <- model$residuals

plot(fitted(model),residuals)
abline(0,0)

```
## Residual vs. Predictor
```{r, }
res1 <- model$residuals
plot(residuals, x_house_val)
```

```{r, fig.cap = "Both median house value and median income are distributed roughly normal."}
# Normality
qqnorm(x_house_val)
qqline(x_house_val)
qqnorm(y_inc)
qqline(y_inc)
```



# Plotting Variables
```{r, fig.cap='This graph displays the linear relationship between the median value of a house and the median income of the people living in a particular block in California. We can observe that this relationship is close when in low income area and with cheaper but this relationship is more varied as the median price of houses and people\'s income rises.'}

model <- lm(y_inc ~ x_house_val, data = housing)
p <- ggplot(housing, aes(x = x_house_val, y = y_inc)) + 
  geom_point() + 
  labs(x = 'Median house value of a California block in USD',
       y = 'Median house income in USD') +xlim(0, (max(x_house_val+ 20000)))

p + geom_smooth(method = 'lm',
  formula = 'y ~ x',
  se = F)
```

## Confidence Interval
```{r}
model <- lm(y_inc ~ x_house_val, data = housing)
model_summary <- summary(model)
beta_hat_0 <- model$coefficients[[1]]
beta_hat_1 <- model$coefficients[[2]]

t_0 <- model_summary$coefficients["x_house_val", "t value"]
p_value <- model_summary$coefficients["x_house_val" , "Pr(>|t|)"]
se <- model_summary$coefficients["x_house_val" , "Std. Error"]

lower <- beta_hat_1 - t_quantile * se_beta_hat_1
upper <- beta_hat_1 + t_quantile * se_beta_hat_1
```
With 95% confidence level, the estimated true median house value of a housing block in California 1990s is said to be between 0.0000107653771071449  and 0.000012634788259628.


```{r}
model <- lm(y_inc ~ median_house_value, data = housing)
model_summary <- summary(model)
beta_hat_0 <- model$coefficients[[1]]
beta_hat_1 <- model$coefficients[[2]]

model_summary

# t_0 <- model_summary$coefficients["x_house_val", "t value"]
# p_value <- model_summary$coefficients["x_house_val" , "Pr(>|t|)"]
# se <- model_summary$coefficients["x_house_val" , "Std. Error"]

lower <- beta_hat_1 - t_quantile * se_beta_hat_1
upper <- beta_hat_1 + t_quantile * se_beta_hat_1

```

# More Confidence Interval
```{r}
# 1 Random Sample House to Predict.
house1 <- sample(1:length(housing$households), 1,replace = FALSE)
house1 <- housing[house1, ]

fittemp <- lm(median_income ~ median_house_value, data = housing)

mean_house_val <- mean(housing$median_house_val)

# point estimate for mean responses
c(1, as.numeric(mean_house_val)) %*% coef(fittemp)

# c(y0 <- sum(x0 * coef(model)))
# point estimate for specific response
# c(1, as.numeric(house1$median_house_value)) %*% coef(fittemp)
```


# CI prediction
```{r}
model <- lm(median_income ~ median_house_value, data = housing)

x <- model.matrix(model) #fitted model to matrix
x0 <- apply(x, 2, mean) # mean median house value

# predict(model, new = data.frame(t(x0)), interval = "confidence")
# 
predict(model, new = data.frame(t(x0)), interval = "prediction")

c(y0 <- sum(x0 * coef(model)))

```
The predicted median house value of a typical block in California in the 1990s is $\$(3.851572 * 10,000)$ or $\$38,500$




