---
title: "California Housing Price in 1990"
author:
  - "Phuc Lu"
  - "Meghna Chandrasekar"
  - "Sophia Li"
  - "Youngju Kwon"
output:
  pdf_document:
    extra_dependencies: ["float"]
    latex_engine: xelatex
  html_document: default
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: preamble-latex.tex
spacing: single
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning=FALSE, message=FALSE}
library(readr) # So we can read csv file
library(skimr)
# library(mapview)
# library(sf)
library(dplyr)
library(ggplot2)
options(scipen=999) # Undo Scientific notation
library(tidyr)

```

[Source](https://www.kaggle.com/datasets/camnugent/california-housing-prices/data) from Kaggle.

```{r}
set.seed(1)
housing_base <- read_csv('housing.csv',show_col_types = FALSE)
housing <- sample(1:length(housing_base$households), 600,replace = FALSE)
housing <- housing_base[housing, ]

## Converting median housing age into categorical data
## Categories: new house, moderately old house, new house
## Intervals 20 years apart from 0 - 60

age = housing$housing_median_age

for (i in 1:length(age)){
    if (age[i] <= 20){
    housing$housing_median_age[[i]] <- "NEW"
  }
  else if (age[i] > 20 && age[i] <= 40){
    housing$housing_median_age[[i]] <- "MODERATE"
  }
  else if (age[i] > 40 && age[i] <= 60){
    housing$housing_median_age[[i]]<-"OLD"
  }
}
```

```{r}
# Scaling Unit of Median Income to its correct values
inc <- housing$median_income

# Income to its correct values
inc <- housing$median_income
for (i in 1:length(inc)){
  if (inc[i] < inc[i]*10000){
  housing$median_income[[i]] = inc[i]*10000
  }
}
```

```{r, eval = FALSE}
housing_sf <- st_as_sf(housing, coords = c("longitude", "latitude"), crs = 4326)
# mapview(housing_sf)
```

```{r}
# Median House Value and Median Income

# Median House Value within the block
#  measured in US Dollars

# Median income of people in the block
# Measured in tens of thousand of US dollars

# Blocks are defined by its longitude and latitude

```



## Our Simple Linear Regression Model is given by
$$
(\text{Median Income})_i = \beta_0 + \beta_1\text{(Median House Value)}_i + \epsilon_i
$$
where $\beta_0$ and $\beta_1$ are unknown constants and $\epsilon_i$ is the random error.

## Hypothesis
We want to know if there is a linear relationship between the median value of houses in a particular block in California in the 1990s and the people living in the block's median income. In other words, we want to test to see if the value of people's house has any linear relationship to their income.\
$$
H_0: \beta_1 = 0 \text{ vs. } H_a: \beta_1 \neq 0
$$

## Plotting Variables
```{r, fig.pos='H', fig.width = 5, fig.height = 3, fig.cap='This graph displays the linear relationship between the median value of a house and the median income of the people living in a particular block in California. We can observe that this relationship is close when in low income area and with cheaper but this relationship is more varied as the median price of houses and people\'s income rises.'}

model <- lm(median_income ~ median_house_value, data = housing)
p <- ggplot(housing, aes(x = median_house_value, y = median_income)) + 
  geom_point() + 
  labs(x = 'Median house value of a block (USD)',
       y = 'Median house income (USD)') +xlim(0, (max(housing$median_house_value+ 20000))) + ggtitle ("Linear Model of Median Income and Median House Value")

p + geom_smooth(method = 'lm',
  formula = 'y ~ x',
  se = F) + theme(plot.title = element_text(size=12))
```

## Checking Assumptions
```{r, fig.pos='H',fig.align='center', fig.width = 7, fig.height = 4, fig.cap= "These graphs show the fan shape which means that as we move go from cheaper to more expensive houses, the variation increases. This fails the assumption of constant variance. As a consequence, we can expect the estimate of the variance, or MSE to be overestimate or underestimate some values of the predictors."}
# Residual Vs. Fitted Plot
model <- lm(median_income ~ median_house_value, data = housing)
house_val <- housing$median_house_value
inc <- housing$median_income

residuals <- model$residuals
par(mfrow=c(1, 2))

plot(fitted(model),residuals, 
     pch = 19,
     main = "Residual vs. Fitted Model",
     xlab = "Residuals",
     ylab = "Fitted Model",
     cex.main = 1)
abline(0,0)

res1 <- model$residuals
plot(residuals, house_val, 
     pch = 19,
     main = "Residual vs. Predictor",
     xlab = "Residuals",
     ylab = "Median House Value",
     cex.main = 1)

```

## Normality
```{r,fig.pos='H',fig.align='center', fig.width = 7, fig.height = 4, fig.cap = "The median house value is distributed approximately Normal"}

par(mfrow = c(1,2))
qqnorm(house_val, pch = 19,
       main = "Normaly Q-Q Plot of Median House Value",
       cex.main = .8)
qqline(house_val)

qqnorm(inc, pch = 19,
       main = "Normaly Q-Q Plot of Median Income",
       cex.main = 1
)
qqline(inc)
```

# Hypothesis Testing
```{r}
## Manual Calculation

alpha <- 0.05

house_val <- housing$median_house_value
inc <- housing$median_income

if (length(house_val) == length(inc)){
  n <- length(house_val)
}else{
  print(paste("Inconsistent n"))
}

# Sxy and Sxx
Sxy <- sum((house_val - mean(house_val)) * (inc - mean(inc)))
Sxx <- sum((house_val - mean(house_val))^2)

# beta 1 and beta 0
beta_hat_1 <- Sxy/Sxx
beta_hat_0 <- mean(inc) - (beta_hat_1 * mean(house_val))

# prediction and residual
y_hat <- beta_hat_0 + (beta_hat_1 * house_val)
residuals <- inc - y_hat

# Mean squared residuals
SSres <- sum(residuals^2)
ms_res <- SSres/ (n - 2)

# Standard Error of beta 1
se_beta_hat_1 <- sqrt(ms_res / Sxx)

# t-statistic
t0 <- beta_hat_1 / se_beta_hat_1

t_quantile <- qt(p = 1 - alpha/2, df = n - 2)

# p-value
p_value <- 2 * pt(abs(t0), df = n - 2, lower.tail = FALSE)

# list(beta_hat_0 = beta_hat_0, beta_hat_1 = beta_hat_1, t0 = t0, t_quantile = t_quantile)
```

```{r}
# print(paste('Sxx: ', Sxx))
# print(paste('beta hat 0: ', beta_hat_0))
# print(paste('beta hat 1: ', beta_hat_1))
# print(paste('sigma hat^2: ', ms_res_house))
# print(paste('t0: ', t0))
# print(paste('t-distribution: ', t_quantile))
# print(paste('p_value: ', p_value))
# print(paste('standard error: ', se_beta_hat_1))
```

Reject $H_0$ if $|t_0| > t_{0.0025, 598}$\
We found that $t_0 = 24.5834068785213$\
So, $|24.5834068785213| >  1.9639388980555$\

```{r}
# print(paste('At the 0.05 significant level, we fail to reject H0. This means that given a particular block in California, there is a linear relation between the median house value and the median income of the people living in the block.'))
```
# Decision Rule:
At the 0.05 significant level, we reject H0. There is a linear relationship between the median house value and the median income of the people according to the California Census data in the 1990s.\

## Confidence Interval
```{r}
model <- lm(median_income ~ median_house_value, data = housing)
model_summary <- summary(model)

beta_hat_0 <- model$coefficients[[1]]
beta_hat_1 <- model$coefficients[[2]]

t_0 <- model_summary$coefficients["median_house_value", "t value"]
p_value <- model_summary$coefficients["median_house_value" , "Pr(>|t|)"]
se <- model_summary$coefficients["median_house_value" , "Std. Error"]
se <- model_summary$coefficients["median_house_value", "Std. Error"]

lower <- beta_hat_1 - t_quantile * se
upper <- beta_hat_1 + t_quantile * se

ci_med_house_val <- confint(model, 'median_house_value', level = 0.95)
```

With 95% confidence level, the true mean median value of houses the 1990s in California is estimated to be between \$0.1076538 and \$0.1263479.\

# More Confidence Interval
```{r}
set.seed(1)
# 1 Random Sample House to Predict.
rand_block <- sample(1:length(housing$households), 1,replace = FALSE)
rand_block <- housing[rand_block, ]

fittemp <- lm(median_income ~ median_house_value, data = housing)
skimmed_housing <- skim(housing)

# Building a Mean Data Frame
mean_skimmed_house <- skimmed_housing %>%
  select(skim_variable, numeric.mean)

mean_df <- mean_skimmed_house %>%
  pivot_wider(names_from = skim_variable, values_from = numeric.mean)


# point estimate for mean responses
point_est_mean_resp <- c(1, as.numeric(mean_df$median_income)) %*% coef(fittemp)

# CI for specific response
mean_median_income <- predict(fittemp, newdata = mean_df, interval = 'confidence', level = 0.95)


# point estimate for specific response
point_est_spec_resp <- c(1, as.numeric(rand_block$median_income)) %*% coef(fittemp)

# PI for specific response
pred_median_income <- predict(fittemp, newdata = rand_block, interval = 'prediction', level = 0.95)
```
### Confidence Interval
With 95% confidence, the mean median income of California residents with measurements equal to the average in the data in the 1990s is estimated to be between \$37431.57 and \$39599.88

### Prediction Interval
We took a random sample of one block from the data and got a block with coordinates 32.72 latitude and -117.13 longitude.

With 95% confidence, the predicted median income of California residents in the 1990s within this block is estimated to be between \$1827.977 and \$55009.23.

#  Conclusion
- We learned that the the $\hat\beta_1\text{(median house value)}$ is 0.1170008, which is small but not neglectible. This means that there is a small association between value of someone's house how much they make.

- Our confidence interval for the median house value doesn't make sense despite imputing the correct data. We get that people were more poorer in the 1990s compare to modern times, but houses should not cost a few cents to buy.

- When we plot our simple linear model, we observe that the data for median income cuts off at \$500,000 with 21 observations. This is unexplained by the source. We believe that this might've resulted from flaws in data collection process. The collector probably rounded off larger and less common values down to \$500,000. However, we also to know if there are data cleaning methods that can improve this issue.