---
title: "Project Step 1"
author: "Phuc, Meghna, Sophia, Youngju"
output:
  html_document: default
  pdf_document: default
date: "2024-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r}
library(readr)
library(MASS)
```

```{r}
# Stroke Prediction Data Subset
set.seed(1)
stroke_base_ds <- read_csv('healthcare-dataset-stroke-data.csv',show_col_types = FALSE)
random_sample <- sample(1:length(stroke_base_ds$id), 600,replace = FALSE)
stroke_ds <- stroke_base_ds[random_sample, ]
```

```{r}
names(stroke_ds)
```
## What Each Attribute means
1) id: unique identifier of each patient.
2) gender: "Male", "Female" or "Other"
3) age: age of the patient
4) hypertension: 0 if the patient doesn't have hypertension, 1 if the patient has hypertension
5) heart_disease: 0 if the patient doesn't have any heart diseases, 1 if the patient has a heart disease
6) ever_married: "No" or "Yes"
7) work_type: "children", "Govt_jov", "Never_worked", "Private" or "Self-employed"
8) Residence_type: "Rural" or "Urban"
9) avg_glucose_level: average glucose level in blood
10) bmi: body mass index
11) smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown"*
12) stroke: 1 if the patient had a stroke or 0 if not
*Note: "Unknown" in smoking_status means that the information is unavailable for this patient


```{r}
# summary(stroke_ds$age)
head(stroke_ds)
```

```{r}
sample_size <- length(stroke_ds$id)
sample_size
```

### How Old Are Our Subjects in the Data?

```{r}
age_data <- stroke_ds$age

max_age <- max(age_data)
min_age <- min(age_data)
# summary(age_data)
bin_width <- (max_age - min_age) / 10
age_ranges <- c("0-10", "11-20", "21-30", "31-40", "41-50", "51-60", "61-70", "71-80", "81-90", "91-100")

hist(age_data,
     main = "Histogram of Patient's Ages",
     xlab = "Patient's Age",
     ylab = "Number of Patients",
     col = "skyblue",
     xlim = c(0, 100),
     ylim = c(0, max(age_data) + 50),
     breaks = seq(0, 100, 10),
     labels = age_ranges,
     cex.axis = 0.4
     )
```

<hr>

### What is the Ratio Between Male and Female Patients In Our Data?
```{r}
data_gender <- table(stroke_ds$gender)
num_males <- sum(stroke_ds$gender == "Male")
num_females <- sum(stroke_ds$gender == "Female")

pie(data_gender, 
    col = c("chartreuse3", "brown1"),
    main = "Ratio between number of Male and Female Patients",
    clockwise = TRUE
    # labels = c(num_males, num_females)
    )

# print(paste("There are", num_males, "male patients and", num_females, "female patients", "in our dataset."))
# print(paste("The ratio between the number of male patients and female patients is", as.fractions(num_males/num_females)))
# print(paste("This means that there are more females patients in our data than male patients."))
```
There are 252 male patients and 348 female patients in our dataset.
The ratio is 21:29.
This means that there are more female patients than male patients.

### Could our dataset be biased?
```{r}
tb_gender <- table(stroke_base_ds$gender)
num_males_base <- sum(stroke_base_ds$gender == "Male")
num_females_base <- sum(stroke_base_ds$gender == "Female")
total <- num_males_base + num_females_base

print(paste("Percent of Men in population", num_males/total * 100, '%' ))
print(paste("Percent of Women in population", num_females/total * 100, '%'))

as.fractions(num_males_base/num_females_base)

```


### The types of occupation in our data set
```{r}
occupation_data <- table(stroke_ds$work_type)
occupation_data
occupation_bp <- barplot(occupation_data, 
        main = "The Types of Occupation of Patients in Dataset", 
        xlab = "Type of Occupations",
        ylab = "Number of Patients",
        col = "skyblue",
        ylim =c(0, max(occupation_data) + 200),
        names.arg = c("Children", "Government Job", "Never Worked", "Private", "Self Employed"),
        cex.names = 0.7,
        las = 1,
        )
par(mar = c(5, 5, 5, 2)) # Bottom Left Top Right

text(x = occupation_bp, y = occupation_data + 0.1, labels = occupation_data, pos = 3)
```


### Hypertension (high blood pressure)
##### When the pressure in the patient's blood vessels are too high (140/90 mmHg or higher).\newline

```{r}
table_hypertension <- table(stroke_ds$hypertension)
# stroke_ds$hypertension
barplot_hypertension <- barplot(table_hypertension,
        main = "Number of Patients with Hypertension in Data", 
        xlab = "Patient Response", 
        ylab = "Number of Patients",
        col = "skyblue",
        names.arg = c('No', 'Yes'),
        xlim = c(0,3),
        ylim = c(0, max(table_hypertension) + 100))

text(x = barplot_hypertension, y = table_hypertension + 0.1, labels = table_hypertension, pos = 3)
```


## When Do Strokes Start Appear in our Patients?
### Map the probability of stroke ~ age.
```{r}
# past_stroke <- stroke_ds$stroke
# past_stroke
# barplot(age_data ~ past_stroke)

```


